Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`

### 2021-03-18 manuscript.Rmd
fixed missing text

In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`. Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented. `bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011c; @danecek2021a]. This implementation has the advantage of being very fast. `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012a; @howtofa].

In this work, `bcftools` and `freebayes` variant callers were used to obtain biallelic SNPs and results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.

Best method could depend on nature of the sequencing data: eg. whole-genome, RAD-seq, RNA-seq, scRNA-seq, etc, and the depth of coverage.

@@

@@

```{r load_data, message=FALSE}
#! {R code}
library(tidyverse, quietly = T)

# Overlap counts table; add
Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
  distinct() %>% 
  bind_rows(tibble(SNPs = c('mpileup', 'freebayes'), 
                   Count = c(.[1,2][[1]] + .[3,2][[1]], 
                             .[2,2][[1]] + .[3,2][[1]]))) %>%  
  arrange(desc(Count))
```

```{r load_data2, message=FALSE, echo=FALSE}
# Allele frequencies
cnames <- c('chrom','pos', 'n_alleles','n_chr','a1','a2')
AlleleFreq <-  
  bind_rows(
    read_tsv('./data/mpileup.filter.frq', col_names = cnames, skip = 1L) %>% 
      mutate(caller = 'mpileup'),
    read_tsv('./data/freebayes.filter.frq', col_names = cnames, skip = 1L) %>% 
      mutate(caller = 'freebayes')
  ) %>% 
  mutate(allele = str_extract(a1, '[A-Z]+'), 
         frequency = str_remove_all(a1, '[A-Z]+:'),
         allele2 = str_extract(a2, '[A-Z]+'), 
         frequency2 = str_remove_all(a2, '[A-Z]+:')
  ) %>% 
  mutate(across(contains('freq'), as.numeric)) %>%
  mutate(
    maf = ifelse(frequency > 0.5, 1 - frequency, frequency),
    minor_allele = ifelse(maf == frequency, allele, allele2)
  )

@@

vcf_df <- SiteDepth %>%
  rename(chrom = chr) %>% 
  left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))

cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
qual_df <- bind_rows(
  read_tsv("./data/mpileup.quality.stats",
                 skip = 1L, col_names = cnames) %>% 
  mutate(caller = "mpileup"),
  read_tsv("./data/freebayes.quality.stats",
                 skip = 1L, col_names = cnames) %>% 
  mutate(caller = "freebayes")
  ) %>% 
  mutate(caller = as_factor(caller))

@@

```{r figs, echo=F}
## this chunk contains code for all the plots
library(ggbeeswarm)
library(patchwork)
library(rcartocolor)
library(gridExtra)
library(grid)

# Allele frequency histogram (alleles stacked)
fig_allelefreq <- 
  ggplot(AlleleFreq) +
  geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
  facet_wrap(caller~., scales = 'free_y') +
  labs(x = 'minor allele frequency', y='loci') +
  theme_bw()

# The distribution of depth per-site summed across individuals
fig_sitedepth <- 
  ggplot(SiteDepth, aes(x = allele_depth, fill = caller, color = caller)) +
  geom_density(alpha = 0.2) +
  geom_vline(aes(xintercept = mean(allele_depth), color = caller),
             lty = 2) +
  scale_x_log10() +
  labs(x = 'depth per locus\n(sum of 10 individuals)', y = 'density') +
  scale_color_carto_d(palette = 2, direction = -1) +
  scale_fill_carto_d(palette = 2, direction = -1) +
  guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
  theme_bw() +
  theme(legend.position = c(.8,.7))

@@

# Allele depth over maf, coloured by n chromosomes
fig_alleledepth_over_maf <- 
  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_y_log10() +
  facet_wrap(~caller) +
  scale_color_viridis_c() +
  labs(x = "Minor allele frequency",
       y = "Allele depth", 
       color="Number of\nchromosomes") +
  theme_bw()

# Quality score distribution by caller fig
fig_quality_dist <- qual_df %>% 
  mutate(QUAL = round(quality/25)*25) %>% 
  group_by(QUAL, caller) %>% 
  summarise(n_snps = sum(n_snps),
            ts  = sum(ts),
            tv = sum(tv)) %>% 
  pivot_longer(cols = c(ts,tv)) %>% 
  ggplot(aes(x = QUAL, y = value, fill = name)) +
  geom_col(width = 19) +
  facet_wrap(~caller) +
  scale_fill_carto_d(palette = 2) +
  labs(x="quality score", y = "loci", fill = NULL) +
  scale_y_log10() +
  theme_bw()

@@

```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=3.5, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted. (B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}

@@

@@

```{r fig_allele_maf_nchr, echo=F, fig.align='center', fig.height=4, fig.cap="Allele depth and minor allele frequency do not show any correlation; minor allele frequency is constrained to specific values by the number of chromosomes represented"}
fig_alleledepth_over_maf
```

## Mean depth by subject

Mean depth for each individual across all loci was computed as follows:

We can see the all individuals have greater mean depth at SNPs called by `freebayes` than `bcftools mpileup` (fig. 4)

```{r fig_indv_depth, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual is greater for SNPs called by  freebayes than bcftools."}
fig_indvdepth
```

### Quality scores

I obtained quality scores for each loci from bctools stats

```{r engine = 'bash', eval=FALSE}
## quality scores...
bcftools stats mpileup.filter.recode.vcf > mpileup.filter.recode.stats
grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats bcftools stats freebayes.filter.recode.vcf > freebayes.filter.recode.stats
grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats
```

```{r fig3, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Minor allele fre"}
fig_alleledepth_over_maf
```

```{r fig4, echo=F, fig.height = 3, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="The quality score distributions of freebayes and bcftools mpileup "}
fig_quality_dist
```

@@

```{=html}
<!-- Zeggini, E., & Morris, A. (2015). Assessing Rare Variation in Complex Traits Design and Analysis of Genetic Studies (1st ed. 2015.). Springer 
New York. <https://doi.org/10.1007/978-1-4939-2824-8> -->
```

>>>>

In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`. Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented. `bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011, @danecek2021]. This implementation has the advantage of being very fast. `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012, @howtof].

In this work, I used `bcftools` and `freebayes`callers to find biallelic SNPs. The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies. Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015], though others have reported high concordance and accuracy in their results [@hwang2015]. Best method could depend on nature of the sequencing data: eg. whole-genome, RAD-seq, RNA-seq, scRNA-seq, etc, and the depth of coverage.

@@

-   Mean depth for each individual across all loci

@@

```{r load data, message=FALSE}
#! {R code}
library(tidyverse, quietly = T)

# Overlap counts table; add
Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
  distinct() %>% 
  bind_rows(
    tibble(SNPs = c('mpileup', 'freebayes'), 
           Count = c(.[1,2][[1]] + .[3,2][[1]], 
                     .[2,2][[1]] + .[3,2][[1]]))) %>%  
  arrange(desc(Count))

# Allele frequencies
cnames <- c('chrom','pos', 'n_alleles','n_chr','a1','a2')
AlleleFreq <-  
  bind_rows(
    read_tsv('./data/freebayes.filter.frq', col_names = cnames, skip = 1L) %>% 
      mutate(caller = 'freebayes'),
    read_tsv('./data/mpileup.filter.frq', col_names = cnames, skip = 1L) %>% 
      mutate(caller = 'mpileup')
  ) %>% 
  mutate(allele = str_extract(a1, '[A-Z]+'), 
         frequency = str_remove_all(a1, '[A-Z]+:'),
         allele2 = str_extract(a2, '[A-Z]+'), 
         frequency2 = str_remove_all(a2, '[A-Z]+:')
  ) %>% 
  mutate(across(contains('freq'), as.numeric)) %>%
  mutate(
    maf = ifelse(frequency > 0.5, 1 - frequency, frequency),
    minor_allele = ifelse(maf == frequency, allele, allele2)
  )

@@

vcf_df <- SiteDepth %>%
  rename(chrom = chr) %>% 
  left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))
# glimpse(vcf_df)

cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
qual_df <- 
  read_tsv("./data/freebayes.quality.stats",
                 skip = 1L, col_names = cnames) %>% 
  mutate(caller = "freebayes") %>% 
  bind_rows(
    read_tsv("./data/mpileup.quality.stats",
                 skip = 1L, col_names = cnames) %>% 
  mutate(caller = "mpileup")
  )

@@

```{r figs, echo=F}
library(ggbeeswarm)
library(patchwork)
library(rcartocolor)
library(gridExtra)
library(grid)
# Allele frequency histogram (alleles stacked)
fig_allelefreq <- 
  ggplot(AlleleFreq) +
  geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
  facet_wrap(caller~., scales = 'free_y') +
  labs(x = 'minor allele frequency', y='loci') +
  theme_bw()

# The distribution of depth per-site summed across individuals
fig_sitedepth <- 
  ggplot(SiteDepth, aes(x = allele_depth, fill = caller, color = caller)) +
  geom_density(alpha = 0.12) +
  geom_vline(aes(xintercept = mean(allele_depth), color = caller),
             lty = 2) +
  scale_x_log10() +
  labs(x = 'depth per locus', y = 'density') +
  guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
  theme_bw() +
  theme(legend.position = c(.8,.7))

@@

# Allele depth over maf, coloured by n chromosomes
fig_alleledepth_over_maf <- 
  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_y_log10() +
  facet_wrap(~caller) +
  scale_color_viridis_c() +
  labs(x = "Minor allele frequency",
       y = "Allele depth", 
       color="Number of\nchromosomes") +
  theme_bw()
```

@@

```{r fig1, echo=F, fig.align = 'center', fig.width=7, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted. (B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}

@@

```{=html}
<!-- Zeggini, E., & Morris, A. (2015). Assessing Rare Variation in Complex Traits Design and Analysis of Genetic Studies (1st ed. 2015.). Springer 
New York. <https://doi.org/10.1007/978-1-4939-2824-8> -->
```
```{r engine = 'bash', eval=FALSE}
## quality scores...
grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats 
grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats 
```

```{r echo = F}
fig_quality_dist <- qual_df %>% 
  mutate(QUAL = round(quality/25)*25) %>% 
  group_by(QUAL, caller) %>% 
  summarise(n_snps = sum(n_snps),
            ts  = sum(ts),
            tv = sum(tv)) %>% 
  pivot_longer(cols = c(ts,tv)) %>% 
  ggplot(aes(x = QUAL, y = value, fill = name)) +
  geom_col(width = 19) +
  facet_wrap(~caller) +
  scale_fill_carto_d(palette = 2) +
  labs(x="quality score", y = "loci", fill = NULL) +
  scale_y_log10() +
  theme_bw()

@@

```{r}
fig_alleledepth_over_maf
```

```{r fig2, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
fig_indvdepth
```

```{r fig3, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
fig_alleledepth_over_maf
```

```{r}

@@
